\documentclass[sigconf, nonacm]{acmart}

% ============================================================================
% PRISMA SUMMER SCHOOL 2026 -- GROUP WORK REPORT
% ============================================================================
\setcopyright{none}
\settopmatter{printacmref=false, printccs=false, printfolios=true}
\acmDOI{}
\acmISBN{}
\acmConference[PRISMA '26]{PRISMA Summer School on Uncertainty in Energy Systems}{June 15--19, 2026}{University of Geneva, Switzerland}

% ============================================================================
% PACKAGES
% ============================================================================
\usepackage{amsmath,amsfonts}
\usepackage{booktabs}
\usepackage{siunitx}
\usepackage{graphicx}
\usepackage{float}

\sisetup{per-mode=symbol, exponent-product=\times, group-separator={,}}

% ============================================================================
\begin{document}
% ============================================================================

\title{Hindcast Evaluation of the SCI Ensemble: Is the Probability of Net Zero by 2070 Biased?}

\author{Lucas Alvarez}
\email{lualpi@upv.es}
\affiliation{%
  \institution{Universitat Polit\`ecnica de Val\`encia}
  \city{Valencia}
  \country{Spain}
}

\author{Matteo Catania}
\email{matteo.catania@polimi.it}
\affiliation{%
  \institution{Politecnico di Milano}
  \city{Milan}
  \country{Italy}
}

\author{R\'emi Paccou}
\email{remi.paccou@proton.me}
\affiliation{%
  \institution{CIRED}
  \city{Nogent-sur-Marne}
  \country{France}
}
\affiliation{%
  \institution{Schneider Electric}
  \city{Grenoble}
  \country{France}
}

% ----------------------------------------------------------------------------
\begin{abstract}
The Scenario Compass Initiative (SCI) assembles 1,564 integrated assessment model pathways, of which 497 reach global net zero CO$_2$ by 2070, suggesting a naive probability of 32\%. But not all scenarios are equally credible. We evaluate the ensemble against observed data over 2010--2025 for CO$_2$ emissions, coal, solar PV, and wind capacity. We find that the ensemble systematically underestimates 2025 emissions and coal consumption, and that net-zero scenarios carry significantly larger hindcast errors than non-net-zero ones. Correcting for ensemble imbalance across model families preserves this finding. Economy-integrated models show a persistent optimistic bias on decarbonisation, while energy-system models are better centred but miss macro shocks. These results suggest that the naive 32\% overstates the probability of reaching net zero by 2070.
\end{abstract}

\keywords{hindcast evaluation, scenario ensemble, net zero, forecast error, integrated assessment models}

\maketitle

% ============================================================================
\section{Introduction}\label{sec:intro}
% ============================================================================

The Scenario Compass Initiative (SCI 2025) assembles 1,564 pathways from 65 integrated assessment models (IAMs). Of these, 497 reach global net zero CO$_2$ emissions by 2070, yielding a naive estimate $P(\text{NZ2070}) = 497/1564 \approx 32\%$. This estimate treats every scenario as equally credible. Yet these scenarios were built at different times, with different base-year calibrations, and encode different assumptions about technology deployment and policy ambition. Some have already been contradicted by observed data.

We ask: \emph{does correcting for historical accuracy change the probability of reaching net zero by 2070?} To answer this, we compare scenario projections against observed data over 2010--2025, measure forecast errors, and test whether net-zero scenarios are systematically less accurate than others. If they are, the naive 32\% is biased upward.

% ============================================================================
\section{Data and Methods}\label{sec:methods}
% ============================================================================

\paragraph{Scenario data.} We use the SCI 2025 v1.0 global ensemble in IAMC format, covering 2010--2100 in 5-year steps. The meta-sheet provides the year of net zero CO$_2$ for each scenario; we flag those with net zero $\leq 2070$ ($|\mathcal{N}| = 497$).

\paragraph{Observed data.} We compile global observations at the SCI time steps ($t \in \{2010, 2015, 2020, 2025\}$) from three sources: CO$_2$ emissions from the Global Carbon Budget 2025, coal primary energy from the Energy Institute Statistical Review 2025, and solar PV and wind capacity from IRENA Renewable Capacity Statistics 2026.

\paragraph{Error metrics.} For each scenario $j$, variable $v$, and year $t$:
\begin{equation}
  \varepsilon_{j,v,t} = y_{v,t} - \hat{y}_{j,v,t}
\end{equation}
where $y$ is the observed value and $\hat{y}$ is the projection. A positive $\varepsilon$ indicates the scenario underestimates the observed value. We compute the Mean Error (ME~$= \overline{\varepsilon}$, capturing bias), Mean Absolute Error (MAE~$= \overline{|\varepsilon|}$, capturing typical error magnitude), and Root Mean Square Error (RMSE~$= \sqrt{\overline{\varepsilon^2}}$, penalising large deviations). Each metric is computed in three views: averaged over scenarios (one value per year), averaged over time (one value per scenario), and over both (a single summary number).

\paragraph{Ensemble weighting.} The raw ensemble is unbalanced: five model families (REMIND, MESSAGE, IMAGE, WITCH, POLES) contribute 66\% of scenarios. We weight each scenario by $1/n_f$, where $n_f$ is the number of scenarios from its model family, so that each of the 17 families receives equal total weight. Weights are recomputed within the NZ2070 and non-NZ subsets to preserve a fair comparison.

\paragraph{Model classification.} We classify families by economic structure: \emph{economy} models with endogenous macro-economy (CGE: AIM, GEM-E3, IMACLIM, EPPA, CGEM; Ramsey growth: REMIND, WITCH, MERGE; 8 families, 701 scenarios) and \emph{energy-system} models with exogenous GDP (MESSAGE, IMAGE, POLES, COFFEE, GCAM, TIAM, PROMETHEUS; 7 families, 883 scenarios).

% ============================================================================
\section{Results}\label{sec:results}
% ============================================================================

\paragraph{The ensemble undershoots recent trends.}
Table~\ref{tab:view1} shows family-weighted error metrics by year for CO$_2$ emissions. The ensemble is well calibrated in 2010--2015 (ME close to zero). In 2020, models overshoot by $-1{,}572$ Mt~CO$_2$, missing the COVID-induced dip. In 2025, models undershoot by $+2{,}582$ Mt~CO$_2$: the post-COVID rebound exceeded what most scenarios anticipated. Only 32\% of scenarios project coal within $\pm 10\%$ of the observed 2025 value of 164~EJ. Coal reached a record high in 2024--2025, while the ensemble median projected 140~EJ.

\begin{table}[h]
\centering
\caption{Family-weighted forecast errors for CO$_2$ emissions (Mt~CO$_2$/yr).}
\label{tab:view1}
\begin{tabular}{lrrr}
\toprule
Year & ME & MAE & RMSE \\
\midrule
2010 & $+405$ & $750$ & $970$ \\
2015 & $+15$ & $732$ & $1{,}023$ \\
2020 & $-1{,}572$ & $1{,}860$ & $2{,}433$ \\
2025 & $+2{,}582$ & $3{,}550$ & $4{,}537$ \\
\bottomrule
\end{tabular}
\end{table}

\paragraph{Net-zero scenarios are more wrong.}
Table~\ref{tab:view2} compares per-scenario errors (averaged over time) between the NZ2070 and non-NZ groups. NZ scenarios carry a positive ME of $+607$, meaning they systematically project less CO$_2$ than observed --- i.e., they assume faster decarbonisation than has occurred. Non-NZ scenarios are nearly unbiased (ME~$= -23$). The MAE gap ($2{,}165$ vs $1{,}746$) confirms that NZ scenarios are not only biased but also less accurate in magnitude.

\begin{table}[h]
\centering
\caption{Family-weighted per-scenario errors for CO$_2$ (Mt~CO$_2$/yr).}
\label{tab:view2}
\begin{tabular}{lrrrr}
\toprule
Group & ME$_j$ & MAE$_j$ & RMSE$_j$ & $n$ \\
\midrule
All & $+321$ & $1{,}770$ & $2{,}354$ & 1{,}591 \\
NZ2070 & $+607$ & $2{,}165$ & $2{,}976$ & 497 \\
non-NZ & $-23$ & $1{,}746$ & $2{,}270$ & 1{,}094 \\
\bottomrule
\end{tabular}
\end{table}

\paragraph{Economy models are more optimistic.}
Splitting by model class, economy models carry a persistent positive bias (ME~$= +613$), projecting lower emissions than observed across all years. Energy-system models are nearly centred (ME~$= -41$). However, energy models miss the 2020 COVID shock more severely ($-2{,}164$ vs $-1{,}264$), consistent with their bottom-up structure that tracks technical trajectories but not macro shocks. In 2025, both classes undershoot comparably ($+2{,}541$ vs $+2{,}717$): the post-COVID rebound is a shared blind spot.

% ============================================================================
\section{Discussion and Conclusions}\label{sec:discussion}
% ============================================================================

Our hindcast evaluation reveals a structural pattern: the SCI ensemble collectively underestimates the persistence of fossil fuels and overestimates the pace of near-term decarbonisation. This bias is concentrated in the net-zero scenarios, which assumed faster coal phase-out and emission reductions than what has materialised through 2025. The finding is robust to ensemble rebalancing across model families.

This has a direct implication for the guiding question of this exercise. If scenarios that reach net zero by 2070 are the ones that most poorly track recent reality, then the naive estimate $P(\text{NZ2070}) \approx 32\%$ is likely biased upward. Filtering the ensemble for historical credibility --- retaining only scenarios within a given error threshold on CO$_2$ and coal --- would reduce this probability, though the exact magnitude depends on the filtering criterion.

Two caveats are important. First, the 2020--2025 period includes an unprecedented exogenous shock (COVID-19 and its rebound) that no scenario was designed to capture. Some of the measured error is attributable to this shock rather than to structural model deficiencies. Second, hindcast accuracy over a 15-year window does not guarantee forecast accuracy over 45 years: a model that correctly tracks coal in 2025 may still be wrong about 2070.

% ============================================================================
\bibliographystyle{ACM-Reference-Format}
\begin{thebibliography}{10}

\bibitem{gcb2025} Friedlingstein, P. et al. (2026). Global Carbon Budget 2025. \emph{Earth System Science Data}, 18.

\bibitem{iea2026} IEA (2026). \emph{Global Energy Review 2026}. Paris: International Energy Agency.

\bibitem{ei2025} Energy Institute (2025). \emph{Statistical Review of World Energy 2025}. London.

\bibitem{irena2026} IRENA (2026). \emph{Renewable Capacity Statistics 2026}. Abu Dhabi.

\bibitem{sci} Scenario Compass Initiative, \url{https://scenariocompass.org}.

\end{thebibliography}

\end{document}
